SwiftKey Capstone Proj. Milestone Report

### 2019-11-15 draft1.Rmd
some progress

Probs of words completing OBSERVED trigrams
```{r}
gamma2 <- 0.5  # bigram discount
gamma3 <- 0.5  # trigram discount
bigPre <- 'and_tell'

@@

obs_trigs <- getObsTrigs(bigPre, trigs)  # get trigrams and counts
# convert counts to probabilities
qbo_obs_trigrams <- getObsTriProbs(obs_trigs, bigrs, bigPre, gamma3)
qbo_obs_trigrams
```

Find UNOBSERVED trigram tail words
```{r}
## Returns a character vector which are the tail words of unobserved trigrams
## that start with the first two words of obsTrigs (aka the bigram prefix).
## These are the words w in the set B(w_i-2, w_i-1) as defined in the section
## describing the details of equation 17.
##
## obsTrigs - character vector of observed trigrams delimited by _ of the form:
##            w3_w2_w1 where w3_w2 is the bigram prefix
## unigs - 2 column data.frame of all the unigrams in the corpus:
##         ngram = unigram
##         freq = frequency/count of each unigram
getUODTTails = function(obsTrigs, unigs) {

}

@@

unobs_trig_tails <- getUnobsTrigTails(obs_trigs$ngram, unigs)
head(unobs_trig_tails, 5)
```

Calculate discounted probability mass at bigram lvl
```{r}
## Returns the total probability mass discounted from all observed bigrams
## calculated from equation 14.  This is the amount of probability mass which
## is redistributed to UNOBSERVED bigrams. If no bigrams starting with
## unigram$ngram[1] exist, 0 is returned.
##
## unigram - single row, 2 column frequency table. The first column: ngram,
##           contains the w_i-1 unigram (2nd word of the bigram prefix). The
##           second column: freq, contains the frequency/count of this unigram.
## bigrams - 2 column data.frame or data.table. The first column: ngram,
##           contains all the bigrams in the corpus. The second column:
##           freq, contains the frequency or count of each bigram.
## bigDisc - amount to discount observed bigrams
getAlphaBigram <- function(unigram, bigrams, bigDisc=0.5) {
    # get all bigrams that start with unigram
    regex <- sprintf("%s%s%s", "^", unigram$ngram[1], "_")
    bigsThatStartWithUnig <- bigrams[grep(regex, bigrams$ngram),]
    if(nrow(bigsThatStartWithUnig) < 1) return(0)
    alphaBi <- 1 - (sum(bigsThatStartWithUnig$freq - bigDisc) / unigram$freq)

@@

unig <- str_split(bigPre, "_")[[1]][2]
unig <- unigs[unigs$ngram == unig,]
alpha_big <- getAlphaBigram(unig, bigrs, gamma2)
alpha_big
```

Calculate backed-off probabilities q_bo for bigrams
```{r}
## Returns a character vector of backed off bigrams of the form w2_w1. These 
## are all the (w_i-1, w) bigrams where w_i-1 is the tail word of the bigram
## prefix bigPre and w are the tail words of unobserved bigrams that start with
## w_i-1.
##
## bigPre - single-element char array of the form w2_w1 which are first two
##          words of the trigram we are predicting the tail word of
## unobsTrigTails - character vector that are tail words of unobserved trigrams
getBoBigrams <- function(bigPre, unobsTrigTails) {
    w_i_minus1 <- str_split(bigPre, "_")[[1]][2]
    boBigrams <- paste(w_i_minus1, unobsTrigTails, sep = "_")
    return(boBigrams)
}

## Returns a two column data.frame of backed-off bigrams in the first column
## named ngram and their frequency/counts in the second column named freq.
## 
## bigPre -  single-element char array of the form w2_w1 which are first two
##           words of the trigram we are predicting the tail word of
## unobsTrigTails - character vector that are tail words of unobserved trigrams
## bigrs - 2 column data.frame or data.table. The first column: ngram,
##         contains all the bigrams in the corpus. The second column:
##         freq, contains the frequency/count of each bigram.
getObsBoBigrams <- function(bigPre, unobsTrigTails, bigrs) {
    boBigrams <- getBoBigrams(bigPre, unobsTrigTails)
    obs_bo_bigrams <- bigrs[bigrs$ngram %in% boBigrams, ]
    return(obs_bo_bigrams)
}

## Returns a character vector of backed-off bigrams which are unobserved.
##
## bigPre -  single-element char array of the form w2_w1 which are first two
##           words of the trigram we are predicting the tail word of
## unobsTrigTails - character vector that are tail words of unobserved trigrams
## obsBoBigram - data.frame which contains the observed bigrams in a column
##               named ngram
getUnobsBoBigrams <- function(bigPre, unobsTrigTails, obsBoBigram) {
    boBigrams <- getBoBigrams(bigPre, unobsTrigTails)
    unobs_bigs <- boBigrams[!(boBigrams %in% obsBoBigram$ngram)]
    return(unobs_bigs)
}

## Returns a dataframe of 2 columns: ngram and probs.  Values in the ngram
## column are bigrams of the form: w2_w1 which are observed as the last
## two words in unobserved trigrams.  The values in the prob column are
## q_bo(w1 | w2) calculated from from equation 10.
##
## obsBoBigrams - a dataframe with 2 columns: ngram and freq. The ngram column
##                contains bigrams of the form w1_w2 which are observed bigrams
##                that are the last 2 words of unobserved trigrams (i.e. "backed
##                off" bigrams). The freq column contains integers that are
##                the counts of these observed bigrams in the corpus.
## unigs - 2 column data.frame of all the unigrams in the corpus:
##         ngram = unigram
##         freq = frequency/count of each unigram
## bigDisc - amount to discount observed bigrams
getObsBigProbs <- function(obsBoBigrams, unigs, bigDisc=0.5) {
    first_words <- str_split_fixed(obsBoBigrams$ngram, "_", 2)[, 1]
    first_word_freqs <- unigs[unigs$ngram %in% first_words, ]
    obsBigProbs <- (obsBoBigrams$freq - bigDisc) / first_word_freqs$freq
    obsBigProbs <- data.frame(ngram=obsBoBigrams$ngram, prob=obsBigProbs)

@@

## Returns a dataframe of 2 columns: ngram and prob.  Values in the ngram
## column are unobserved bigrams of the form: w2_w1.  The values in the prob
## column are the backed off probability estimates q_bo(w1 | w2) calculated
## from from equation 16.
##
## unobsBoBigrams - character vector of unobserved backed off bigrams
## unigs - 2 column data.frame of all the unigrams in the corpus:
##         ngram = unigram
##         freq = frequency/count of each unigram
## alphaBig - total discounted probability mass at the bigram level
getQboUnobsBigrams <- function(unobsBoBigrams, unigs, alphaBig) {
    # get the unobserved bigram tails
    qboUnobsBigs <- str_split_fixed(unobsBoBigrams, "_", 2)[, 2]
    w_in_Aw_iminus1 <- unigs[!(unigs$ngram %in% qboUnobsBigs), ]
    # convert to data.frame with counts
    qboUnobsBigs <- unigs[unigs$ngram %in% qboUnobsBigs, ]
    denom <- sum(qboUnobsBigs$freq)
    # converts counts to probabilities
    qboUnobsBigs <- data.frame(ngram=unobsBoBigrams,
                               prob=(alphaBig * qboUnobsBigs$freq / denom))

@@

>>>>

Probs of words completing OBSERVED trigrams
```{r}
gamma2 <- 0.5  # bigram discount
gamma3 <- 0.5  # trigram discount
bigPre <- 'is_book'

@@

obs_trigs <- getODTrigs(bigPre, trigs)  # get trigrams and counts
# convert counts to probabilities
qbo_obs_trigrams <- getODTProbs(obs_trigs, bigrs, bigPre, gamma3)
qbo_obs_trigrams
```

Find UNOBSERVED trigram tail words
```{r}
## Returns a character vector which are the tail words of unobserved trigrams
## that start with the first two words of obsTrigs (aka the bigram prefix).
## These are the words w in the set B(w_i-2, w_i-1) as defined in the section
## describing the details of equation 17.
##
## obsTrigs - character vector of observed trigrams delimited by _ of the form:
##            w3_w2_w1 where w3_w2 is the bigram prefix
## unigs - 2 column data.frame of all the unigrams in the corpus:
##         ngram = unigram
##         freq = frequency/count of each unigram
getUODTTails = function(obsTrigs, unigs) {
    tails = str_split_fixed(obsTrigs, "_", 3)[, 3]
    unobs_tails = unigs[!(unigs$ngram %in% tails), ]$ngram
    return(unobs_tails)
}

@@

unobs_trig_tails <- getUODTTails(obs_trigs$ngram, unigs)
head(unobs_trig_tails, 10)
```

Calculate discounted probability mass at bigram lvl
```{r}
## Returns the total probability mass discounted from all observed bigrams
## calculated from equation 14.  This is the amount of probability mass which
## is redistributed to UNOBSERVED bigrams. If no bigrams starting with
## unigram$ngram[1] exist, 0 is returned.
##
## unigram - single row, 2 column frequency table. The first column: ngram,
##           contains the w_i-1 unigram (2nd word of the bigram prefix). The
##           second column: freq, contains the frequency/count of this unigram.
## bigrams - 2 column data.frame or data.table. The first column: ngram,
##           contains all the bigrams in the corpus. The second column:
##           freq, contains the frequency or count of each bigram.
## bigDisc - amount to discount observed bigrams
getAlphaBi = function(unigram, bigrams, bigDisc = 0.5) {
    regex = sprintf("%s%s%s", "^", unigram$ngram[1], "_")
    bgrs_with_unig = bigrams[grep(regex, bigrams$ngram), ]
    if(nrow(bgrs_with_unig) < 1) return(0)
    alphaBi = 1 - (sum(bgrs_with_unig$freq - bigDisc) / unigram$freq)
    return(alphaBi)
}
#returns 0 if no observed bigrams

#mike's code
getAlphaBigram <- function(unigram, bigrams, bigDisc=0.5) {
    # get all bigrams that start with unigram
    regex <- sprintf("%s%s%s", "^", unigram$ngram[1], "_")
    bigsThatStartWithUnig <- bigrams[grep(regex, bigrams$ngram),]
    if(nrow(bigsThatStartWithUnig) < 1) return(0)
    alphaBi <- 1 - (sum(bigsThatStartWithUnig$freq - bigDisc) / unigram$freq)

@@

unig <- str_split(bigPre, "_")[[1]][2]
unig <- unigs[unigs$ngram == unig,]
alpha_big <- getAlphaBi(unig, bigrs, gamma2)
alpha_big
```

Calculate backed-off probabilities q_bo for bigrams
```{r}
## Returns a character vector of backed off bigrams of the form w2_w1. 
## These are all the (w_i-1, w) bigrams where w_i-1 is the tail word of the bigram
## prefix bigPre and w are the tail words of unobserved bigrams that start with
## w_i-1.
##
## bigPre - single-element char array of the form w2_w1 which are first two
##          words of the trigram we are predicting the tail word of
## unobsTrigTails - character vector that are tail words of unobserved trigrams
getBOBigrs = function(bigPre, UODTrigTails) {
    w_i_1 = str_split(bigPre, "_")[[1]][2]
    boBigrams = paste(w_i_1, UODTrigTails, sep = "_")
    return(boBigrams)
}

#mike's code
getBoBigrams <- function(bigPre, unobsTrigTails) {
    w_i_minus1 <- str_split(bigPre, "_")[[1]][2]
    boBigrams <- paste(w_i_minus1, unobsTrigTails, sep = "_")
    return(boBigrams)
}

## Returns a two column data.frame of backed-off bigrams in the first column
## named ngram and their frequency/counts in the second column named freq.
## 
## bigPre -  single-element char array of the form w2_w1 which are first two
##           words of the trigram we are predicting the tail word of
## unobsTrigTails - character vector that are tail words of unobserved trigrams
## bigrs - 2 column data.frame or data.table. The first column: ngram,
##         contains all the bigrams in the corpus. The second column:
##         freq, contains the frequency/count of each bigram.
getODBOBigrs = function(bigPre, UODTrigTails, bigrs) {
    boBigrams = getBOBigrs(bigPre, UODTrigTails)
    obs_bo_bigrs = bigrs[bigrs$ngram %in% boBigrams, ]
    return(obs_bo_bigrs)
}

#mike's code
getObsBoBigrams <- function(bigPre, unobsTrigTails, bigrs) {
    boBigrams <- getBoBigrams(bigPre, unobsTrigTails)
    obs_bo_bigrams <- bigrs[bigrs$ngram %in% boBigrams, ]
    return(obs_bo_bigrams)
}

## Returns a character vector of backed-off bigrams which are unobserved.
##
## bigPre -  single-element char array of the form w2_w1 which are first two
##           words of the trigram we are predicting the tail word of
## unobsTrigTails - character vector that are tail words of unobserved trigrams
## obsBoBigram - data.frame which contains the observed bigrams in a column
##               named ngram
getUODBOBigrs = function(bigPre, UODTrigTails, ODBOBigrs) {
    boBigrs = getBOBigrs(bigPre, UODTrigTails)
    unobs_bigrs = boBigrs[!(boBigrs %in% ODBOBigrs$ngram)]
    return(unobs_bigrs)
}

#mike's code
getUnobsBoBigrams <- function(bigPre, unobsTrigTails, obsBoBigram) {
    boBigrams <- getBoBigrams(bigPre, unobsTrigTails)
    unobs_bigs <- boBigrams[!(boBigrams %in% obsBoBigram$ngram)]
    return(unobs_bigs)
}

## Returns a dataframe of 2 columns: ngram and probs.  Values in the ngram
## column are bigrams of the form: w2_w1 which are observed as the last
## two words in unobserved trigrams.  The values in the prob column are
## q_bo(w1 | w2) calculated from from equation 10.
##
## obsBoBigrams - a dataframe with 2 columns: ngram and freq. The ngram column
##                contains bigrams of the form w1_w2 which are observed bigrams
##                that are the last 2 words of unobserved trigrams (i.e. "backed
##                off" bigrams). The freq column contains integers that are
##                the counts of these observed bigrams in the corpus.
## unigs - 2 column data.frame of all the unigrams in the corpus:
##         ngram = unigram
##         freq = frequency/count of each unigram
## bigDisc - amount to discount observed bigrams
getODBigrsProbs = function(ODBOBigrs, unigs, bigDisc = 0.5) {
    first_words = str_split_fixed(ODBOBigrs$ngram, "_", 2)[ ,1]
    first_word_freqs = unigs[unigs$ngram %in% first_words, ]
    ODBigrsProbs = (ODBOBigrs$freq - bigDisc) / first_word_freqs$freq
    ODBigrsProbs = data.frame(ngram = ODBOBigrs$ngram, prob = ODBigrsProbs)

    return(ODBigrsProbs)
}

#mike's code
getObsBigProbs <- function(obsBoBigrams, unigs, bigDisc=0.5) {
    first_words <- str_split_fixed(obsBoBigrams$ngram, "_", 2)[, 1]
    first_word_freqs <- unigs[unigs$ngram %in% first_words, ]
    obsBigProbs <- (obsBoBigrams$freq - bigDisc) / first_word_freqs$freq
    obsBigProbs <- data.frame(ngram=obsBoBigrams$ngram, prob=obsBigProbs)

@@

## Returns a dataframe of 2 columns: ngram and prob.  Values in the ngram
## column are unobserved bigrams of the form: w2_w1.  The values in the prob
## column are the backed off probability estimates q_bo(w1 | w2) calculated
## from from equation 16.
##
## unobsBoBigrams - character vector of unobserved backed off bigrams
## unigs - 2 column data.frame of all the unigrams in the corpus:
##         ngram = unigram
##         freq = frequency/count of each unigram
## alphaBig - total discounted probability mass at the bigram level
# CHECK
getUODBOBigrsProbs = function(UODBOBigrs, unigs, alphaBig) {
    #unobserved bigram tails
    UODBigrsProbs = str_split_fixed(UODBOBigrs, "_", 2)[, 2]
    w_Aw_i_1 = unigs[!(unigs$ngram %in% UODBigrsProbs), ]
    #dataframe with counts
    UODBigrsProbs = unigs[unigs$ngram %in% UODBigrsProbs, ]
    qD = sum(UODBigrsProbs$freq)
    #probabilities
    UODBigrsProbs = data.frame(ngram = UODBOBigrs, 
                               prob = (alphaBig * UODBigrsProbs$freq / qD))
    return(UODBigrsProbs)
}

#mike's code
getQboUnobsBigrams <- function(unobsBoBigrams, unigs, alphaBig) {
    # get the unobserved bigram tails
    qboUnobsBigs <- str_split_fixed(unobsBoBigrams, "_", 2)[, 2]
    w_in_Aw_iminus1 <- unigs[!(unigs$ngram %in% qboUnobsBigs), ]
    # convert to data.frame with counts
    # PROBLEM HERE, w_in_Aw_iminus1 is not used to get qboUnobsBigs
    qboUnobsBigs <- unigs[unigs$ngram %in% qboUnobsBigs, ]
    denom <- sum(qboUnobsBigs$freq)
    # converts counts to probabilities
    qboUnobsBigs <- data.frame(ngram=unobsBoBigrams,
                               prob=(alphaBig * qboUnobsBigs$freq / denom))

@@

#get backed off bigrams
bo_bigrs = getBOBigrs(bigPre, unobs_trig_tails)
#separate observed bigrams from unobserved bigrams
obs_bo_bigrs = getODBOBigrs(bigPre, unobs_trig_tails, bigrs)
unobs_bo_bigrs = getUODBOBigrs(bigPre, unobs_trig_tails, obs_bo_bigrs)
#qbo
qbo_obs_bigrams = getODBigrsProbs(obs_bo_bigrs, unigs, gamma2)

unig <- str_split(bigPre, "_")[[1]][2]
unig <- unigs[unigs$ngram == unig,]
#qbo redistributed
qbo_unobs_bigrams = getUODBOBigrsProbs(unobs_bo_bigrs, unigs, alpha_big)
